Assessment of zonally symmetric and asymmetric components of the Southern Annular Mode using a novel approach

### 2021-04-27 analysis/paper/asymsam.Rmd
Cambia a line wraping por oración.

Most authors describe the SAM as a zonally symmetric pattern, a fact that is
reflected not only in its name, but also in the various methods used to
characterise it. Of the several different indices presented in the literature,
many of them are based on zonal means of sea level pressure or geopotential
height [@ho2012]. @gong1999 defined the SAM index as the zonal mean sea level
pressure difference between 40\degree S and 60\degree S, which is also the
definition used by the station-based index in @marshall2003. @baldwin2009
proposed defining the Northern and Annular modes as the leading EOF of the
zonally averaged geopotential height at each level.

@@

The impact of the zonally asymmetric component of the SAM at regional scales has
not been studied in detail yet. The positive phase of the SAM is associated with
colder-than-normal temperatures over Antarctica and warmer-than-normal
temperatures at higher latitudes [@jones2019] (and vice versa for negative SAM).
But there are significant deviations from this zonal mean response, notably in
the Antarctic Peninsula and the South Atlantic [@fogt2012]. The SAM-related
signal on precipitation anomalies behaves similarly, although with even greater
deviation from zonal symmetry [@lim2016]. The importance of zonal asymmetries of
the SAM on these impacts have been studied in certain regions. For example, the
SAM-precipitation relationship in Southeastern South America can be explained by
the Pacific-South American (PSA)-like zonally asymmetric circulation associated
with the SAM [@silvestri2009; @rosso2018]. @fan2007 also found that
precipitation in East Asia was impacted by the variability of only the Western
Hemisphere part of the SAM.

@@

```{r compute\nobreakdash-SAM}
SAM <- hgt[lat <= -20] %>% 
  .[lev == 700] %>%
  .[, hgt := hgt_a*sqrt(cos(lat*pi/180))] %>% 
  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1, data = .SD))), 
    by = .(lev)]

@@

@@

By performing the EOF analysis using data for all months we are assuming that
the zonally asymmetric structure of the SAM is the same at all seasons. The
latter was assessed by computing geopotential height zonal anomalies by
projecting the first EOF of each season independently. The following seasons
were considered -- December to February (DJF), March to May (MAM), June to
August (JJA) and September to November (SON). Results are very similar to each
other in the troposphere (Figure \@ref(fig:A5), row 1) and show spatial
correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON). In the
stratosphere (Figure \@ref(fig:A5), row 2), patterns are similar for all seasons
except DJF, when the wave-1 zonal anomalies are rotated 90\degree in comparison
with the rest of the year. Spatial correlations in the stratosphere are between
-0.24 (DJF with SON) and 0.95 (MAM with JJA). Therefore, the results confirm
that the zonal asymmetric structure of the SAM is very similar throughout most
of the year with the exception of DJF in the stratosphere.

@@

```{r compute-asymsam}
indices_file <- data_path("derived", "indices.Rds")
if (file.exists(indices_file)) {
  indices <- readRDS(indices_file)
  indices[, `:=`(estimate = -estimate, 
                 estimate_norm = -estimate_norm)]
} else {
  stop("Run 02-compute-eofs.R")
}

@@

```{r test-aao}
# test that my indices hace the correct sign. The Full index needs to be **positively** correlated
# with the AAO index.
aao <- rsoi::download_aao(TRUE, data_path("raw", "aao.csv")) %>% 
  as.data.table() %>% 
  .[, .(time = lubridate::as_datetime(Date), aao = AAO)]

cor_aao <- indices[term == "full" & lev == 700] %>% 
  .[aao, on = .NATURAL] %>% 
  na.omit() %>% 
  .[, cor(estimate, aao)]

@@

Figure \@ref(fig:cross-correlation) shows (zero-lag) cross-correlation across
levels for the SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices. For the SAM
(Figure \@ref(fig:cross-correlation)a), high values below 100 hPa reflect the
vertical (zero-lag) coherency throughout the troposphere. Above 100 hPa,
correlation between levels does falls off more rapidly, indicating less coherent
(zero-lag) variability. But correlations between tropospheric and
lower-to-middle stratospheric levels are still relatively high (e.g. grater than
0.4 between tropospheric levels and levels below 30 hPa). A\nobreakdash-SAM and
S\nobreakdash-SAM (Figure \@ref(fig:cross-correlation)b and c, respectively)
share similar high level of coherency in the troposphere but they differ in
their stratospheric behaviour. Stratospheric coherency is stronger for the
A\nobreakdash-SAM than the S\nobreakdash-SAM. The stratospheric
S\nobreakdash-SAM seems to connect more strongly to the troposphere than the
stratospheric A\nobreakdash-SAM.

@@

In the stratosphere, the spatial pattern associated with the SAM is clearly
dominated by a zonally symmetric, monopolar structure (Figure
\@ref(fig:2d-regr)a.1) which is not centred in the South Pole. On the other
hand, the monopole associated with S\nobreakdash-SAM (Figure
\@ref(fig:2d-regr)c.1) is more symmetric. Furthermore, the regression pattern of
A\nobreakdash-SAM is characterized by a wave-1 structure with centres over the
Bellinghausen Sea on the Western Hemisphere and Davis Sea in the Eastern
Hemisphere

@@

```{r wave-amplitude, fig.cap = "(ref:wave-amplitude-caption)", fig.width=4, fig.height=3, fig.env = "figure*", cache = FALSE}
indices_2d_yearly %>% 
  .[lat < -20] %>%
  .[lev %in% c(50, 700)] %>% 
  .[, FitWave(estimate, 0:3), by = .(lat, lev, term)] %>% 
  ggplot(aes(lat, amplitude)) +
  geom_line(aes(color = factor(k))) +
  scale_color_brewer("Wavenumber", palette = "Dark2") +
  scale_y_continuous("Amplitude") +
  scale_x_latitude(ticks = 20) +
  coord_flip() +
  facet_grid(lev~term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
  tag_facets("cr", position = "tr")  +
  no_grid 
```

@@

The wave-3 pattern from Figure \@ref(fig:2d-regr)b.2 is very similar to the PSA
Pattern [@mo1987; @kidson1988] which is a teleconnection pattern associated with
the ENSO [@karoly1989]. Indeed, @fogt2011 showed that there is a significant
relationship between the SAM and the ENSO. The correlation between the SAM and
the ENSO as measured by the Oceanic Niño Index [ONI, @bamston1997] is
`r cor_texts[term == "full"]$cor` (p-value
`r cor_texts[term == "full"]$p.value`). This relationship is captured mainly by
the A\nobreakdash-SAM, as this index has a partial correlation of
`r enso_sam_partial[index == "oni" & term == "asym"]$partial_correlation`
(p-value `r enso_sam_partial[index == "oni" & term == "asym"]$p.value`) with the
ONI, consistent with @fan2007's results, whereas the S\nobreakdash-SAM's partial
correlation with the ONI is essentially null
(`r enso_sam_partial[index == "oni" & term == "sym"]$partial_correlation`;
p-value `r enso_sam_partial[index == "oni" & term == "sym"]$p.value`). The same
analysis was performed using the Multivariate ENSO Index [@wolter2011] and the
Southern Oscillation Index [@ropelewski1987], obtaining similar results. The
latter allows to conclude that these results do not depend on the ENSO index
used.

@@

@silvestri2009 suggests that precipitation impacts linked to the SAM changed
strongly before and after 1980. In particular, the negative relationship with
precipitation in South America was absent in some areas and switched sign in
others in the earlier period. The correlation between ENSO and SAM is similarly
non-stationary, also changing sign before the 1980s [@fogt2006; @clem2013].
Seeing as both the ENSO\nobreakdash-SAM relationship and most of the
precipitation impacts in South America are captured by the A\nobreakdash-SAM,
the results presented here are most likely period-dependent. The decadal
variations of the A\nobreakdash-SAM should be focus of future studies.

@@

All data used in this paper is freely available from their respective sources.
ERA5 data can be obtained via the Copernicus Climate Data Store
([<https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels-monthly-means>](https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels-monthly-means){.uri}).
CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado,
USA, from their Web site at <https://psl.noaa.gov/data/gridded/data.cmap.html>.
The Oceanic Niño Index is available via NOAA's Climate Prediction Center:
<https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt>

>>>>

Most authors describe the SAM as a zonally symmetric pattern, a fact that is reflected not only in its name, but also in the various methods used to characterise it.
Of the several different indices presented in the literature, many of them are based on zonal means of sea level pressure or geopotential height [@ho2012].
@gong1999 defined the SAM index as the zonal mean sea level pressure difference between 40\degree S and 65\degree S, which is also the definition used by the station-based index in @marshall2003.
@baldwin2009 proposed defining the Northern and Annular modes as the leading EOF of the zonally averaged geopotential height at each level.

@@

The impact of the zonally asymmetric component of the SAM at regional scales has not been studied in detail yet.
The positive phase of the SAM is associated with colder-than-normal temperatures over Antarctica and warmer-than-normal temperatures at lower latitudes [@jones2019] (and vice versa for negative SAM).
But there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic [@fogt2012].
The SAM-related signal on precipitation anomalies behaves similarly, although with even greater deviation from zonal symmetry [@lim2016].
The SAM-precipitation relationship in Southeastern South America can be explained by the Pacific-South American (PSA)-like zonally asymmetric circulation associated with the SAM [@silvestri2009; @rosso2018].
@fan2007 also found that precipitation in East Asia was impacted by the variability of only the Western Hemisphere part of the SAM.

@@

```{r compute-SAM}
set.seed(43)
SAM <- hgt[lat <= -20] %>% 
  .[lev == 700] %>%
  .[, hgt := hgt_a*sqrt(cos(lat*pi/180))] %>% 
  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1, data = .SD))), 
    by = .(lev)]

@@

```{r test-aao1}
# test that my SAM have the correct sign.
aao <- rsoi::download_aao(TRUE, data_path("raw", "aao.csv")) %>% 
  as.data.table() %>% 
  .[, .(time = lubridate::as_datetime(Date), aao = AAO)] %>% 
  .[time %between% as.Date(c("1979-01-01", "2018-12-01"))]

cor_aao <- SAM$eof[[1]]$left %>% 
  .[aao, on = .NATURAL] %>% 
  na.omit() %>% 
  .[, cor(hgt, aao)]

@@

By performing the EOF analysis using data for all months we are assuming that the zonally asymmetric structure of the SAM is the same at all seasons.
The latter was assessed by computing geopotential height zonal anomalies by projecting the first EOF of each season independently.
The following seasons were considered -- December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON).
Results are very similar to each other in the troposphere (Figure \@ref(fig:A5), row 2) and show spatial correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON).
In the stratosphere (Figure \@ref(fig:A5), row 1), patterns are similar for all seasons except DJF, when the wave-1 zonal anomalies are rotated 90\degree in comparison with the rest of the year.
Spatial correlations in the stratosphere are between -0.24 (DJF with SON) and 0.95 (MAM with JJA).
Therefore, the results confirm that the zonal asymmetric structure of the SAM is very similar throughout most of the year.
The only outlier being DJF, which shows much lower correlations with the other seasons at both levels and the weakest zonal anomalies (Figure \@ref{fig:A3}), which is consistent with @fogt2020.
With this in mind, a

@@

```{r compute-asymsam}
indices_file <- data_path("derived", "indices.Rds")
if (file.exists(indices_file)) {
  indices <- readRDS(indices_file)
  indices[, `:=`(estimate = estimate, 
                 estimate_norm = estimate_norm)]
} else {
  stop("Run 02-compute-eofs.R")
}

@@

```{r test-aao2}
# test that my indices hace the correct sign. The Full index needs to be **positively** correlated
# with the AAO index.
cor_aao <- indices[term == "full" & lev == 700] %>% 
  .[aao, on = .NATURAL] %>% 
  na.omit() %>% 
  .[, cor(estimate, aao)]

@@

Figure \@ref(fig:cross-correlation) shows (zero-lag) cross-correlation across levels for the SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices.
For the SAM (Figure \@ref(fig:cross-correlation)a), high values below 100 hPa reflect the vertical (zero-lag) coherency throughout the troposphere.
Above 100 hPa, correlation between levels falls off more rapidly, indicating less coherent (zero-lag) variability.
But correlations between tropospheric and lower-to-middle stratospheric levels are still relatively high (e.g. greater than 0.4 between tropospheric levels and levels below 30 hPa).
A\nobreakdash-SAM and S\nobreakdash-SAM (Figure \@ref(fig:cross-correlation)b and c, respectively) share similar high level of coherency in the troposphere but they differ in their stratospheric behaviour.
Stratospheric coherency is stronger for the A\nobreakdash-SAM than the S\nobreakdash-SAM.
The stratospheric S\nobreakdash-SAM seems to connect more strongly to the troposphere than the stratospheric A\nobreakdash-SAM.

@@

In the stratosphere, the spatial pattern associated with the SAM is clearly dominated by a zonally symmetric, monopolar structure (Figure \@ref(fig:2d-regr)a.1) which is not centred in the South Pole.
On the other hand, the monopole associated with S\nobreakdash-SAM (Figure \@ref(fig:2d-regr)c.1) is more symmetric.
Furthermore, the regression pattern of A\nobreakdash-SAM is characterized by a wave-1 structure with centres over the Drake Passage in the Western Hemisphere and Davis Sea in the Eastern Hemisphere

@@

```{r wave-amplitude, fig.cap = "(ref:wave-amplitude-caption)", fig.width=4, fig.height=3, fig.env = "figure*", cache = FALSE}
indices_2d_yearly %>% 
  .[lat < -20] %>%
  .[lev %in% c(50, 700)] %>% 
  .[, FitWave(estimate, 0:3), by = .(lat, lev, term)] %>% 
  ggplot(aes(lat, amplitude)) +
  geom_line(aes(color = factor(k)), size = 1.2) +
  scale_color_brewer("Wavenumber", palette = "Set1") +
  scale_y_continuous("Amplitude") +
  scale_x_latitude(ticks = 20) +
  coord_flip() +
  facet_grid(lev~term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
  tag_facets("cr", position = "tr")  +
  no_grid 
```

@@

The wave-3 pattern from Figure \@ref(fig:2d-regr)b.2 is very similar to the PSA Pattern [@mo1987; @kidson1988] which is a teleconnection pattern associated with the ENSO [@karoly1989]. Indeed, @fogt2011 showed that there is a significant relationship between the SAM and the ENSO. The correlation between the SAM and the ENSO as measured by the Oceanic Niño Index [ONI, @bamston1997] is `r cor_texts[term == "full"]$cor` (p-value `r cor_texts[term == "full"]$p.value`).
This relationship is captured mainly by the A\nobreakdash-SAM, as this index has a partial correlation of `r enso_sam_partial[index == "oni" & term == "asym"]$partial_correlation` (p-value `r enso_sam_partial[index == "oni" & term == "asym"]$p.value`) with the ONI, consistent with @fan2007's results, whereas the S\nobreakdash-SAM's partial correlation with the ONI is essentially null (`r enso_sam_partial[index == "oni" & term == "sym"]$partial_correlation`; p-value `r enso_sam_partial[index == "oni" & term == "sym"]$p.value`).
The same analysis was performed using the Multivariate ENSO Index [@wolter2011] and the Southern Oscillation Index [@ropelewski1987], obtaining similar results.
The latter allows us to conclude that these results do not depend on the ENSO index used.

@@

@silvestri2009 suggest that precipitation impacts linked to the SAM changed strongly before and after 1980.
In particular, the negative relationship with precipitation in South America was absent in some areas and switched sign in others in the earlier period.
The correlation between ENSO and SAM is similarly non-stationary, also changing sign before the 1980s [@fogt2006; @clem2013].
Seeing as both the ENSO\nobreakdash-SAM relationship and most of the precipitation impacts in South America are captured by the A\nobreakdash-SAM, the results presented here are most likely period-dependent.
The decadal variations of the A\nobreakdash-SAM should be focus of future studies.

@@

All data used in this paper is freely available from their respective sources.
ERA5 data can be obtained via the Copernicus Climate Data Store ([\<https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means\>](https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels-monthly-means){.uri}).
CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at <https://psl.noaa.gov/data/gridded/data.cmap.html>.
The Oceanic Niño Index is available via NOAA's Climate Prediction Center: <https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt>
